PAWpySeed: A parallelized Python/C package for analyzing DFT wavefunctions in the PAW formalism

### 2018-03-06 docs/paper.tex
tie in sbt to the main code
update paper with additional info and revisions

@@

\subsubsection{Projection onto Basis from Different Structure}
In the pure physical picture of a quantum mechanical system, the eigenfunctions of the time
independent Hamiltonian form a complete basis for any function in real space. By expressing
a wavefunction from a system in the basis of the eigenstates of another system, one might
be able to learn about the nature of the wavefunction in question. For example, projecting
the wavefunction of a defect level in a periodic crystal onto the eigenfunctions of the
bulk crystal might yield information about how the defect level is constructed. For example,
the defect state might project primarily onto the conduction band, characterizing it
as a weakly perturbed host state. Such an understanding of defect levels could help select
corrections to perform on the defect system.\ref{eq:1}
\begin{equation}
\centering
\begin{split}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}}=\widetilde{O}+O_M+O_R+O_S+O_N\\
O_M=\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\bra{\widetilde{\phi}_{a\ind _1}}\ket{\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_R=\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_S=\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_N=\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
\end{split}
\label{eq:finp}
\end{equation}
Equation \ref{eq:finp} is derived in Appendix A.

@@

\section*{Appendix A: Derivation of Overlap Operators from PAW Formalism}
The starting point for this derivation is the transformation operator that
is performed on the pseudowavefunction to obtain the all electron wavefunction
\begin{equation}
T=1+\sum_a\sum_l\sum_m\sum_{\epsilon}(\ket{\phi_{a\ind}}-\ket{\widetilde{\phi}_{a\ind}})
\bra{\widetilde{p}_{a\ind}}=1+\sum_i(\ket{\phi_i}-\ket{\widetilde{\phi}_i})
\bra{\widetilde{p}_i}
\label{eq:T}
\end{equation}
where $\phi_{a\ind}$ are all electron (AE) partial waves, $\widetilde{\phi}_{a\ind}$
are pseudo (PS) partial waves, $\widetilde{p}_{a\ind}$ are projector functions,
$a$ are the site indices of each atom in the structure, and $l$, $m$, and $\epsilon$
specify a spherical harmonic and index which uniquely specify partial wave at a given
index. A summation over $i$ (or $j$, as below) represents a summation over $a$, $l$, $m$,
and $\epsilon$. For further details on the PAW method, including the physical significance
and construction of the partial waves and projector functions, see Blochl's original paper
and Kresse and Joubert's paper relating ultrasoft pseudopotentials and PAW.\cite{blochl}
The next step is to define and pseudo operator $\widetilde{A}$ for each operator $A$
such that $\bra{\psi}A\ket{\psi}=\bra{\widetilde{\psi}}\widetilde{A}\ket{\widetilde{\psi}}$.
Because $\ket{\psi}=T\ket{\widetilde{\psi}}$, one can write
\begin{equation}
\widetilde{A}=T^{\dagger}AT
\label{eq:op}
\end{equation}
One can then plug Equation \ref{eq:T} into Equation \ref{eq:op} to find
$$\widetilde{A}=[1+\sum_i\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})]
A[1+\sum_j(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\bra{\widetilde{p}_j}]$$
\begin{equation}
\widetilde{A}=A+\sum_i\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A
+\sum_j A(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\bra{\widetilde{p}_j}
+\sum_i\sum_j\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\bra{\widetilde{p}_j}
\label{eq:res}
\end{equation}
Using the following relation,
$$(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
=\bra{\phi_i}A\ket{\phi_j}-\bra{\widetilde{\phi}_i}A\ket{\phi_j}
-(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A\ket{\widetilde{\phi}_j}$$
$$=\bra{\phi_i}A\ket{\phi_j}-\bra{\widetilde{\phi}_i}A(\ket{\phi_j}-\ket{\widetilde{\phi_j}})
-(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A\ket{\widetilde{\phi}_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}}$$
one can rearrange Equation \ref{eq:res} in the manner of Blochl\cite{blochl}:
\begin{equation}
\begin{split}
\widetilde{A}=A+\sum_i\sum_j[\ket{\widetilde{p}_i}(\bra{\phi_i}A\ket{\phi_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}})\bra{\widetilde{p}_j}] \\
+\sum_i[(1-\sum_j\ket{\widetilde{p}_j}\bra{\widetilde{\phi}_j})A(\ket{\phi_i}-\ket{\widetilde{\phi_i}})
\bra{\widetilde{p}_i}+\ket{\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A(1-\sum_j\ket{\widetilde{\phi}_j}\bra{\widetilde{p}_j})]
\end{split}
\label{eq:nonloc}
\end{equation}
When the operator $A$ is local, then $\sum_j\ket{\widetilde{\phi}_j}\bra{\widetilde{p}_j}=1$,
so the entire second line of Equation \ref{eq:nonloc} vanishes, giving Blochl's formulation:\cite{blochl}
\begin{equation}
\widetilde{A}=A+\sum_i\sum_j\ket{\widetilde{p}_i}(\bra{\phi_i}A\ket{\phi_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}})\bra{\widetilde{p}_j}
\label{eq:loc}
\end{equation}

However, while the overlap operator is local, the simplification in Equation \ref{eq:loc} assumes that
the summations over $i$ and $j$ go over the same species at the same locations in the same lattice;
only the third condition is always satisfied for systems of interest for projections between different
bases. Equation \ref{eq:res} is a more convenient representation than Equation \ref{eq:nonloc} for this
purpose because it is easily seen that all terms between projectors whose augmentation regions do not
overlap vanish. In addition, the POTCAR file in VASP only stores the PS and AE partial waves out to the
edge of the augmentation region, where they are equal but not nonzero, so it is useful to organize the terms
so that they are guaranteed to vanish outside the augmentation regions. Replacing $A$ with unity
allows the overlap between two bands
in different structures $R$ and $S$ with the same lattice to be specified (note that the
summation over $i$ is for structure $R$ and the summation over $j$ is for structure $S$):
\begin{equation}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}}=\widetilde{O}+O_1+O_2+O_3
\label{eq:ov1}
\end{equation}
$$\widetilde{O}=\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
$$O_1=\sum_i\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
$$O_2=\sum_j\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
$$O_3=\sum_i\sum_j\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
The pseudowavefunctions can be expanded as a summation of plane waves,
$$\widetilde{\psi}_{n\mathbf{k}}(\mathbf{r})=\sqrt{\frac{\omega_{\mathbf{k}}}{V}}
e^{i\mathbf{k}\cdot \mathbf{r}}\sum_{\mathbf{G}} C_{n\mathbf{k}\mathbf{G}}
e^{i\mathbf{G}\cdot \mathbf{r}}$$
so the overlap between two pseudowavefunctions per unit cell can be written as
$$\widetilde{O}=\braket{\widetilde{\psi}_{n_1\mathbf{k}_1}|\widetilde{\psi}_{n_2\mathbf{k}_2}}
=\delta_{\mathbf{k}_1,\mathbf{k}_2}
\omega_{\mathbf{k}} \sum_{\mathbf{G}} C^*_{n_1\mathbf{k}_1\mathbf{G}}C_{n_2\mathbf{k}_2\mathbf{G}}$$
In VASP, structures with the same energy, k-points, and lattice will have the same basis set,
so this projection is performed simply by reading plane wave coefficients from the VASP WAVECAR file.

It is important to simplify the calculation of the other terms in equation \ref{eq:ov1} as much
as possible because their calculation can be quite computationally intensive, and the number of
necessary calculations for projecting onto an entire basis set can scale with the number
of sites times the size of the basis set. One major simplification is that if a site $a$ in structure
$R$ and site $b$ in structure $S$ have the same species and position, $a$ and $b$ will only have
overlapping augmentation regions with each other and no other sites. Then, defining $O_{1a}$
as the summation over on-site terms for the identical sites $a$ and $b$ in $O_1$:
$$O_{1a}+O_{2a}+O_{3a}=\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\bra{\widetilde{\phi}_{a\ind _1}}\ket{\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
which is the local operator solution derived by Blochl. All three terms must
be evaluated in full for the other sites, but terms in $O_3$ where $i$ and $j$ correspond
to sites with nonoverlapping augmentation spheres vanish. Therefore, if $M_{RS}$ is the set
of identical sites in the structures $R$ and $S$, $N_R$ and $N_S$ are the sets of sites
in $R$ and $S$ not in $M_{RS}$, and $N_{RS}$ is the set of \emph{pairs} of sites not in
$M$ with overlapping augmentation regions, then
$$
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}}=\widetilde{O}+O_M+O_R+O_S+O_N
$$
$$
O_M=\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\bra{\widetilde{\phi}_{a\ind _1}}\ket{\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_R=\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_S=\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_N=\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
which is Equation \ref{eq:finp}.

>>>>

\title{PAWpySeed: A parallelized Python/C package for analyzing
DFT wavefunctions in the PAW formalism}
\author{Kyle Bystrom\\
        kylebystrom@berkeley.edu\\
		UC Berkeley Department of Materials Science and Engineering}
\date{\today}
\maketitle
\abstract{A parallelized Python and C package has been developed
to perform analysis of density functional theory (DFT) wavefunctions in the
projector augmented wave (PAW) formalism. The primary function of this code
is to calculate the overlap operator of two single-particle Kohn-Sham states
from different periodic structures with the same lattice, which is made nontrivial
by the depency of PAW augmentation regions on the basis of the structure.
The code is potentially useful for the analysis of point defect levels in
solids because it can determine the composition of defect levels in the basis
of the single-particle states of the bulk system.}

@@

\subsubsection{Projection onto Basis from Different Structure}
In the pure physical picture of a quantum mechanical system, the eigenfunctions of the time
independent Hamiltonian form a complete basis for any function in real space. By expressing
a wavefunction from a system in the basis of the eigenstates of another system, one might
be able to learn about the nature of the wavefunction in question. For example, projecting
the wavefunction of a defect level in a periodic crystal onto the eigenfunctions of the
bulk crystal might yield information about how the defect level is constructed. For example,
the defect state might project primarily onto the conduction band, characterizing it
as a weakly perturbed host state. Such an understanding of defect levels could help select
corrections to perform on the defect system.\cite{lany}
\begin{equation}
\centering
\begin{split}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}} & =\widetilde{O}+O_M+O_R+O_S+O_N\\
O_M & =\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\bra{\widetilde{\phi}_{a\ind _1}}\ket{\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_R & =\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_S & =\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_N & =\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
\end{split}
\label{eq:finp}
\end{equation}
Equation \ref{eq:finp} is derived in Appendix A.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|c|}
\hline
Computational Task & $\Theta$ & Frequency*\\
\hline
$\widetilde{O}$ & $BKSW$ & per band\\
$O_M$ & $BKSNP$ & per band\\
$O_R$ and $O_S$ & $BKSNP$ & per band\\
$O_N$ & $BKSNP$ & per band\\
$\braket{\widetilde{p}_i|\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSF\mathrm{log}(F)$ & per structure\\
$(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{n\mathbf{k}}}$ & $BKSNPW$ & per structure\\
spherical Bessel transform partial waves & $NPG\mathrm{log}(G)$ & per structure pair\\
project plane waves onto partial waves & $EPKW$ & per structure pair\\
projections for overlapping aug spheres & $NPL$ & per structure pair\\
\hline
\end{tabular}

\begin{tabular}{c|c}
Symbol & Definition\\
\hline
B & number of bands\\
E & number of elements\\
F & size of FFT grid\\
G & size of partial wave radial grid\\
K & number of k-points\\
L & size of Legendre-Gauss grid around ions\\
N & number of sites**\\
P & number of projector functions\\
S & number of spin states\\
W & number of plane waves\\
\end{tabular}
\caption{Runtime scaling functions for each component of the code and
definitions for shorthand symbols to express runtime.
*The frequency refers to how often the routine is called. "Per band"
indicates that the routine runs once every time a band from one structure
is projected onto all the bands of a basis structure. "Per structure"
indicates that the routine is a "setup" routine that must be called once
for each structure and corresponding band structure to do perform projections.
"Per structure pair" is the same as "per structure" except that the routine
is only run once for a pair of structures with the same lattice and plane-wave
basis set, the bands of one such structure to be used as a basis set for the other.
**Number of sites flexibly refers to the number of sites relevant to the calculation,
which worst-case scales with the total number of sites in the structure. For example,
calculating $O_M$ and $O_N$ only require the sites in sets $M$ and $N_RS$, respectively.}
\label{tab:runtime}
\end{table}

The program is implemented as a Python package that interfaces with C functions through
the Python ctypes package. Time consuming routines are performed in C and parallelized
using openmp, meaning that the code will automatically parallelize on all the resources
of a shared memory computing node. Detailed runtime scaling for each routine is
provided in Table \ref{tab:runtime}. An explanation of how each term is calculated
so as to maximize computational efficiency is summarized in the following subsections.

\subsection{Projection of Pseudowavefunctions onto Partial Waves}

The all electron partial waves fluctuate rapidly near the nucleus,
so the pseudowavefunction cannot be projected onto the partial waves
via a uniform Cartesian grid as is doen with the projector functions.
However, A plane wave can be expanded as a sum of spherical
Bessel functions via Rayleigh expansion.
\begin{equation}
e^{i\mathbf{k} \cdot \mathbf{r}} = 4\pi \sum_{l=0}^{\infty}\sum_{m=-l}^{l}
i^l j_l(kr)Y_l^{m*}(\mathbf{\hat{k}})Y_l^m(\mathbf{\hat{r}})
\label{eq:pwexp}
\end{equation}
Therefore, the overlap operator between a partial wave and plane wave can be evaluated as a summation
of spherical Bessel function overlap integrals. The following definitions are
used in this derivation:

$\mathbf{r_A} = \mathbf{r} - \mathbf{R_A}$, $\phi_i(\mathbf{r})=R_i(r)Y_l^m(\mathbf{\hat{r}})$,
$\widetilde{\phi}_i(\mathbf{r})=\widetilde{R}_i(r)Y_l^m(\mathbf{\hat{r}})$,
$\mathbf{k'}=\mathbf{k}+\mathbf{G}$\\
The derivation goes as follows, with the last line being that needed to 
calculate $O_R$ and $O_S$:

\begin{equation}
\begin{split}
\widetilde{\psi}_{n\mathbf{k}} & =\frac{1}{\sqrt{V}}\sum_{\mathbf{G}}
C_{n\mathbf{k}\mathbf{G}}e^{i\mathbf{k'}\cdot \mathbf{r}}\\
\widetilde{\psi}_{n\mathbf{k}} & =\frac{4\pi}{\sqrt{V}}\sum_{\mathbf{G}}
C_{n\mathbf{k}\mathbf{G}}\sum_{l=0}^{\infty}\sum_{m=-l}^{l}
i^l j_l(k'r_A)Y_l^{m*}(\mathbf{\hat{k'}})Y_l^m(\mathbf{\hat{r}_A})\\
\phi_i(\mathbf{r})\widetilde{\psi}_{n\mathbf{k}} & =\frac{4\pi}{\sqrt{V}}\sum_{\mathbf{G}}
C_{n\mathbf{k}\mathbf{G}}\sum_{l'=0}^{\infty}\sum_{m'=-l'}^{l'}
i^{l'} j_{l'}(k'r_A)Y_{l'}^{m'*}(\mathbf{\hat{k'}})Y_{l'}^{m'}(\mathbf{\hat{r}_A})
Y_l^{m*}(\mathbf{\hat{r}}_A)R_i(r_A)\\
\braket{\phi_i(\mathbf{r})|\widetilde{\psi}_{n\mathbf{k}}} & =
\frac{4\pi i^l}{\sqrt{V}}\sum_{\mathbf{G}} C_{n\mathbf{k}\mathbf{G}}
Y_l^{m*}(\mathbf{\hat{k'}})\int_0^{\infty}drr^2R_i(r_A)j_l(k'r_A)\\
\braket{\phi_i-\widetilde{\phi}_i|\widetilde{\psi}_{n\mathbf{k}}} & =
\frac{4\pi i^l}{\sqrt{V}}\sum_{\mathbf{G}} C_{n\mathbf{k}\mathbf{G}}
Y_l^{m*}(\mathbf{\hat{k'}})\int_0^{\infty}drr^2(R_i(r_A)-\widetilde{R}_i(r_A))j_l(k'r_A)
\end{split}
\end{equation}

After the projection is decomposed into a summation over integrals of spherical
Bessel functions, it can be calculated via spherical Bessel transform. This code
uses the fast spherical Bessel transform method NumSBT developed by Talman.\cite{TALMAN}
It runs in $O(NlnN)$ time and works on a logarithmic grid, making it convenient for use with
the POTCAR representations of the partial waves. The resulting logarithmic k-space
grid can be used to calculate the values of the relevant integrals via interpolation.

\section*{Appendix A: Derivation of Overlap Operators from PAW Formalism}

The following section derives an equation for the overlap operator between one
Kohn-Sham single particle state of one structure R and one Kohn-Sham single
particle state of another structure S, where R and S share a common lattice
and the DFT wavefunctions are constructed with the same plane-wave PAW basis set.
It should be noted this derivation might better be described as a manipulation
of terms of the PAW pseudo-operator, because its main purpose is to present
the overlap operator in a method convenient for computation rather than to derive
an entirely new expression. The derivation also makes particular note of how
the expression of the overlap operator relates to interfacing with the VASP code,
though the expressions are applicable to any PAW code.

The starting point for this derivation is the transformation operator that
is performed on the pseudowavefunction to obtain the all electron wavefunction
\begin{equation}
T=1+\sum_a\sum_l\sum_m\sum_{\epsilon}(\ket{\phi_{a\ind}}-\ket{\widetilde{\phi}_{a\ind}})
\bra{\widetilde{p}_{a\ind}}=1+\sum_i(\ket{\phi_i}-\ket{\widetilde{\phi}_i})
\bra{\widetilde{p}_i}
\label{eq:T}
\end{equation}
where $\phi_{a\ind}$ are all electron (AE) partial waves, $\widetilde{\phi}_{a\ind}$
are pseudo (PS) partial waves, $\widetilde{p}_{a\ind}$ are projector functions,
$a$ are the site indices of each atom in the structure, and $l$, $m$, and $\epsilon$
specify a spherical harmonic and index which uniquely specify partial wave at a given
index. A summation over $i$ (or $j$, as below) represents a summation over $a$, $l$, $m$,
and $\epsilon$. For further details on the PAW method, including the physical significance
and construction of the partial waves and projector functions, see Blochl's original paper
and Kresse and Joubert's paper relating ultrasoft pseudopotentials and PAW.\cite{blochl,kresse_joubert}
The next step is to define a pseudo operator $\widetilde{A}$ for each operator $A$
such that $\bra{\psi}A\ket{\psi}=\bra{\widetilde{\psi}}\widetilde{A}\ket{\widetilde{\psi}}$.
Because $\ket{\psi}=T\ket{\widetilde{\psi}}$, one can write
\begin{equation}
\widetilde{A}=T^{\dagger}AT
\label{eq:op}
\end{equation}
One can then plug Equation \ref{eq:T} into Equation \ref{eq:op} to find
\begin{equation}
\begin{split}
\widetilde{A} = & [1+\sum_i\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})]
A[1+\sum_j(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\bra{\widetilde{p}_j}]\\
\widetilde{A} = & A+\sum_i\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A
+\sum_j A(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\bra{\widetilde{p}_j}\\
& +\sum_i\sum_j\ket{\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\bra{\widetilde{p}_j}
\end{split}
\label{eq:res}
\end{equation}

Using the following relation,
\begin{equation}
\begin{split}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
& =\bra{\phi_i}A\ket{\phi_j}-\bra{\widetilde{\phi}_i}A\ket{\phi_j}
-(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A\ket{\widetilde{\phi}_j}\\
& =\bra{\phi_i}A\ket{\phi_j}-\bra{\widetilde{\phi}_i}A(\ket{\phi_j}-\ket{\widetilde{\phi_j}})
-(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A\ket{\widetilde{\phi}_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}}
\end{split}
\end{equation}

one can rearrange Equation \ref{eq:res} in the manner of Blochl\cite{blochl}:
\begin{equation}
\begin{split}
\widetilde{A}= & A+\sum_i\sum_j[\ket{\widetilde{p}_i}(\bra{\phi_i}A\ket{\phi_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}})\bra{\widetilde{p}_j}] \\
& +\sum_i[(1-\sum_j\ket{\widetilde{p}_j}\bra{\widetilde{\phi}_j})A(\ket{\phi_i}-\ket{\widetilde{\phi_i}})
\bra{\widetilde{p}_i}+\ket{\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})A(1-\sum_j\ket{\widetilde{\phi}_j}\bra{\widetilde{p}_j})]
\end{split}
\label{eq:nonloc}
\end{equation}
When the operator $A$ is local, then $\sum_j\ket{\widetilde{\phi}_j}\bra{\widetilde{p}_j}=1$,
so the entire second line of Equation \ref{eq:nonloc} vanishes, giving Blochl's formulation:\cite{blochl}
\begin{equation}
\widetilde{A}=A+\sum_i\sum_j\ket{\widetilde{p}_i}(\bra{\phi_i}A\ket{\phi_j}
-\bra{\widetilde{\phi_i}}A\ket{\widetilde{\phi_j}})\bra{\widetilde{p}_j}
\label{eq:loc}
\end{equation}

However, while the overlap operator is local, the simplification in Equation \ref{eq:loc} assumes that
the summations over $i$ and $j$ go over the same species at the same locations in the same lattice;
only the third condition is always satisfied for systems of interest for projections between different
bases. Equation \ref{eq:res} is a more convenient representation than Equation \ref{eq:nonloc} for this
purpose because it is easily seen that all terms between projectors whose augmentation regions do not
overlap vanish. In addition, the POTCAR file in VASP only stores the PS and AE partial waves out to the
edge of the augmentation region, where they are equal but not nonzero, so it is useful to organize the terms
so that they are guaranteed to vanish outside the augmentation regions. Replacing $A$ with unity
allows the overlap between two bands
in different structures $R$ and $S$ with the same lattice to be specified (note that the
summation over $i$ is for structure $R$ and the summation over $j$ is for structure $S$):
\begin{equation}
\begin{split}
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}} & =\widetilde{O}+O_1+O_2+O_3\\
\widetilde{O} & =\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_1 & =\sum_i\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_2 & =\sum_j\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}\\
O_3 & =\sum_i\sum_j\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
\end{split}
\label{eq:ov1}
\end{equation}

The pseudowavefunctions can be expanded as a summation of plane waves,
$$\widetilde{\psi}_{n\mathbf{k}}(\mathbf{r})=\sqrt{\frac{\omega_{\mathbf{k}}}{V}}
e^{i\mathbf{k}\cdot \mathbf{r}}\sum_{\mathbf{G}} C_{n\mathbf{k}\mathbf{G}}
e^{i\mathbf{G}\cdot \mathbf{r}}$$
so the overlap between two pseudowavefunctions per unit cell can be written as
$$\widetilde{O}=\braket{\widetilde{\psi}_{n_1\mathbf{k}_1}|\widetilde{\psi}_{n_2\mathbf{k}_2}}
=\delta_{\mathbf{k}_1,\mathbf{k}_2}
\omega_{\mathbf{k}} \sum_{\mathbf{G}} C^*_{n_1\mathbf{k}_1\mathbf{G}}C_{n_2\mathbf{k}_2\mathbf{G}}$$

In VASP, structures with the same energy, k-points, and lattice will have the same basis set,
so this projection is performed simply by reading plane wave coefficients from the VASP WAVECAR file.

It is important to simplify the calculation of the other terms in equation \ref{eq:ov1} as much
as possible because their calculation can be quite computationally intensive, and the number of
necessary calculations for projecting onto an entire basis set can scale with the number
of sites times the size of the basis set. One major simplification is that if a site $a$ in structure
$R$ and site $b$ in structure $S$ have the same species and position, $a$ and $b$ will only have
overlapping augmentation regions with each other and no other sites. Then, defining $O_{1a}$
as the summation over on-site terms for the identical sites $a$ and $b$ in $O_1$:
$$O_{1a}+O_{2a}+O_{3a}=\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\braket{\widetilde{\phi}_{a\ind _1}|\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}$$
which is the local operator solution derived by Blochl. All three terms must
be evaluated in full for the other sites, but terms in $O_3$ where $i$ and $j$ correspond
to sites with nonoverlapping augmentation spheres vanish. Therefore, if $M_{RS}$ is the set
of identical sites in the structures $R$ and $S$, $N_R$ and $N_S$ are the sets of sites
in $R$ and $S$ not in $M_{RS}$, and $N_{RS}$ is the set of \emph{pairs} of sites not in
$M$ with overlapping augmentation regions, then
$$
\braket{\psi_{Rn_1\mathbf{k}}|\psi_{Sn_2\mathbf{k}}}=\widetilde{O}+O_M+O_R+O_S+O_N
$$
$$
O_M=\sum_{a\in M_{RS}}\sum_{l,m}\sum_{\epsilon_1}\sum_{\epsilon_2}
\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_{a\ind _1}}
(\braket{\phi_{a\ind _1}|{\phi_{a\ind _2}}}
-\braket{\widetilde{\phi}_{a\ind _1}|\widetilde{\phi}_{a\ind _2}})
\braket{\widetilde{p}_{a\ind _2}|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_R=\sum_{i\in N_R}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|\widetilde{p}_i}
(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
\ket{\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_S=\sum_{j\in N_S}\bra{\widetilde{\psi}_{Rn_1\mathbf{k}}}(\ket{\phi_j}-\ket{\widetilde{\phi}_j})
\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
$$
O_N=\sum_{i,j\in N_{RS}}\braket{\widetilde{\psi}_{Rn_1\mathbf{k}}|
\widetilde{p}_i}(\bra{\phi_i}-\bra{\widetilde{\phi}_i})
(\ket{\phi_j}-\ket{\widetilde{\phi}_j})\braket{\widetilde{p}_j|\widetilde{\psi}_{Sn_2\mathbf{k}}}
$$
which is Equation \ref{eq:finp}.
